### 2017-06-14 jedm-reviews/round-2/reviewer_b_response.md

> Related works
>
> In prior iterations, the related works did not frame the authors' research
> with respect to similar methods. I think the authors have done an excellent
> job in reworking this section. It is much clearer how the authors' work is
> novel in both the education and data mining/machine learning domains.

Thank you. We greatly appreciate your helpful feedback that helped improve
this aspect of the paper's presentation.

> Abstract and style
>
> The abstract is now sufficiently long and adequately written. The table
> captions have also been fixed. Certain parts of the figures are still hard
> to read, but I understand that the possible fixes here are limited. In that
> regard, the visual issues do not significantly hinder interpretability of
> the concepts.

We again thank you for your help in improving the quality of our abstract.

> Challenges / Limitations
>
> The authors have created a full challenges and limitations section. The
> authors have added a great deal of clarity to the core issues they
> encountered with both the methods and outcomes of their work.

We agree that this added section brings much more clarity to the technical
challenges and potential limitations of our approach, and thank you for the
feedback leading to this improvement to the paper.

> General readability comments:
>
> Previously, I made comments about the readability of the work and the
> authors requested a few examples. I have written down a few quotes from the
> paper along with explanations or fixes for the problems (denoted by an
> arrow ->). It is important to note that these are not all of the issues in
> the paper. I think that authors could still make many improvements to
> increase both the readability and clarity of the work.

Thank you for the detailed list of suggestions. In the revision, we have
incorporated these suggested changes to the wording and have also done
another editing pass to attempt to correct additional errors of similar
form. We hope that the readability and clarity of the work has improved as
a direct result.

### 2017-06-14 sec-intro.tex
Update paper for reviewer b; add response.

The proliferation of massive open online courses (MOOCs) has resulted in a
profound impact on education. As more and more students participate in
these novel educational environments, it is of utmost importance that we be
able to understand the behavioral patterns of students as they interact
with them. While we can easily observe the changes in behavior of students
in real classrooms, MOOCs present a challenge due to their hands-off nature
and sometimes irregular schedule due to being a full-time worker.

At the same time, as more and more learners turn to MOOCs to educate
themselves on various topics, more and more behavioral data is being
collected as part of the system on which the MOOC is offered. This
presents a unique opportunity: the data present in these logs has the power
to aid us in understanding the behavior of students who take our MOOCs,
which is mostly undetectable for instructors of these MOOCs today due to
its vast scale. As a result, the rich data available through these MOOC
logs is highly underutilized today.

What stands in the way? Clearly, intelligent systems to create concise and
digestible summaries of the massive amount of interaction data collected
are needed in order to empower the instructors of these courses. If we can
understand how users are interacting with our MOOCs, we are much more
likely to be able to make changes to these courses that positively impact
learners. We view this paper as attempting to bridge this gap.

How should we represent behavioral patterns, and what does it mean to
understand changes in student behavior with respect to these patterns?
These are still very open questions, and are active areas of
research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
Shih:2010:EDM}. In this paper, we advocate for a particular representation
of student behavior patterns as well as \emph{behavior transitions} that we
believe is simultaneously interpretable but also amenable to unsupervised,
automated discovery via statistical means. Specifically, we choose to
visualize behavior patterns as labeled directed graphs where a node represents a ``behavior state'' (e.g., watching lecture video, or visiting a forum) and a directed edge indicates a transition from one behavior state to another, and node sizes are
proportional to steady-state probabilities and edge widths are proportional
to the probability of leaving a node following that edge. We can use this
same representation for visualizing both the student behavior patterns as
well as the transitioning behavior between them. In
Figure~\ref{fig:motivating-example} we show a hypothetical example of the
kind of output our proposed representation could convey. Here we see two
different behavioral patterns (\ref{fig:motivating-example-0} and
\ref{fig:motivating-example-1}) as well as the \emph{transition behavior}
between these two behavior patterns (\ref{fig:motivating-example-trans}).
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state0.png}
    \caption{Behavior pattern 0\label{fig:motivating-example-0}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state1.png}
    \caption{Behavior pattern 1\label{fig:motivating-example-1}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/trans.png}
    \caption{Transitions between the two behavior patterns to the
    left.\label{fig:motivating-example-trans}}
  \end{subfigure}
  \caption{An idealized example of what our behavior representation could capture.}
  \label{fig:motivating-example}
\end{figure}
Such a visualized state-transition representation is very informative for describing student behavior. Indeed, we could infer many things from even such a simple example: The first might
be that, when students are taking quizzes, they tend to either use the
forum or the videos for support, but not both. They also tend to take
quizzes in a sort of ``cycle'' pattern, indicating perhaps that this course
allows quiz re-takes. Finally, in Figure~\ref{fig:motivating-example-trans}
we could conclude that users tend to change their quiz-taking behavior over
time from one that is more video-focused (pattern 0) to one that is more
forum-focused (pattern 1).

Our goal in this paper is to design a model that can automatically capture
student behavior in this way via unsupervised learning methods applied to
large collections of click logs associated with MOOCs. We view our model
as a component of a system that enables collaboration between the machine
and a human instructor to extract knowledge from these large collections of
MOOC data. Automatically extracting interpretable patterns from the
clickstream data associated with MOOCs is a necessary step in order for
instructors to identify the hidden knowledge in these massive interaction
datasets. Without the availability of a suitable model for identifying
these behavioral patterns, instructors are not empowered to use this
available data to improve their courses without expending extraordinary
amounts of manual effort (even with which the raw data can still be very hard to interpret).

>>>>

The proliferation of massive open online courses (MOOCs) has resulted in a
profound impact on education. As more and more students participate in
these novel educational environments, it is of utmost importance that we be
able to understand the behavioral patterns students exhibit in these
environments. While we can easily observe the changes in behavior of
students in real classrooms, MOOC environments present some significant
challenges in this regard: the structure of the course itself is more
hands-off in nature than that of the traditional classroom (in most cases),
and thus attracts more students that are full-time workers with irregular
learning schedules.

At the same time, this influx of learners turning to MOOC platforms to
educate themselves directly leads to the collection of larger datasets of
behavioral data through the platform's logging. This presents a unique
opportunity: the data present in these logs has the power to aid us in
understanding the behavior of students who take our MOOCs. However, due to
the vast scale of these behavioral logs, student behavior patterns are
mostly undetectable for MOOC instructors and as a result the rich data
available through MOOC logs is highly underutilized today.

What stands in the way? Instructors require intelligent systems to create
concise and digestible summaries of the massive amount of interaction data
collected. If we can understand how users are interacting with our MOOCs,
we are much more likely to be able to make changes to these courses that
positively impact learners. We view this paper as attempting to bridge this
gap.

How should we represent behavioral patterns, and what does it mean to
understand changes in student behavior with respect to these patterns?
These are still very open questions, and are active areas of
research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
Shih:2010:EDM}. In this paper, we advocate for a particular representation
of student behavior patterns as well as \emph{behavior transitions} that we
believe is simultaneously interpretable but also amenable to unsupervised,
automated discovery via statistical means. Specifically, we choose to
visualize behavior patterns as labeled directed graphs where a node
represents a ``behavior state'' (such as watching a lecture video or
visiting a forum), a directed edge indicates a transition from one
behavior state to another, node sizes are proportional to steady-state
probabilities, and edge widths are proportional to the probability of
leaving a node following that edge. We can use this same representation for
visualizing both the student behavior patterns as well as the transitioning
behavior between them. In Figure~\ref{fig:motivating-example} we show a
hypothetical example of the kind of output our proposed representation
could convey.  Here we see two different behavioral patterns
(\ref{fig:motivating-example-0} and \ref{fig:motivating-example-1}) as well
as the \emph{transition behavior} between these two behavior patterns
(\ref{fig:motivating-example-trans}).
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state0.png}
    \caption{Behavior pattern 0\label{fig:motivating-example-0}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state1.png}
    \caption{Behavior pattern 1\label{fig:motivating-example-1}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/trans.png}
    \caption{Transitions between the two behavior patterns to the
    left.\label{fig:motivating-example-trans}}
  \end{subfigure}
  \caption{An idealized example of what our behavior representation could capture.}
  \label{fig:motivating-example}
\end{figure}
Such a visualized state-transition representation is very informative for
describing student behavior. Indeed, we could infer many things from even
such a simple example: The first might be that, when students are taking
quizzes, they tend to either use the forum or the videos for support, but
not both. They also tend to take quizzes in a sort of ``cycle'' pattern,
indicating perhaps that this course allows quiz re-takes. Finally, in
Figure~\ref{fig:motivating-example-trans} we could conclude that users tend
to change their quiz-taking behavior over time from one that is more
video-focused (pattern 0) to one that is more forum-focused (pattern 1).

Our goal in this paper is to design a model that can automatically capture
student behavior in this way via unsupervised learning methods applied to
large collections of click logs associated with MOOCs. We view our model
as a component of a system that enables collaboration between the machine
and a human instructor to extract knowledge from large collections of
MOOC data. Automatically extracting interpretable patterns from the
clickstream data associated with MOOCs is a necessary step in order for
instructors to identify the hidden knowledge in massive interaction
datasets. Without the availability of a suitable model for identifying
behavioral patterns, instructors are not empowered to use this available
data to improve their courses without expending extraordinary amounts of
manual effort (even with which the raw data can still be very hard to
interpret).

### 2017-06-14 sec-limitations.tex
Update paper for reviewer b; add response.

 %irregular behavior
%patterns In particular, because
One potential limitation is that the model is complex and has many parameters,
to truly uncover relevant patterns in the data, thus a large amount of it must
be available to the training algorithm. The assumption that we have a large
amount of sequence data available for training on should generally hold for
most MOOC courses, but this assumption may be problematic if attempting to
apply our model to smaller online (or on-campus) courses.

@@

Our proposed model is flexible in the patterns it can discover in two main
ways. The first is that the granularity of the patterns can be adjusted by
changing the segmentation strategy one uses to divide the user action
stream into discrete ``sessions''. The other is in the number of latent
states $K$ that are used to describe the segmented action sequences. One
drawback of these two lines of flexibility is that there is not necessarily
a clear answer as to the ``correct'' approach for both in any given
scenario. Varying the segmentation strategy changes the granularity of the
patterns that are required to be explained by the latent states, which will
change their meaning. Varying the number of latent states increases the
flexibility of the model, but also may result in latent states that are not
substantially different from the other latent states and/or latent states
with very low probability. This flexibility forces a user of our model to
make some assumptions about the patterns they wish to find (granularity,
diversity), and the model itself does not necessarily provide clear
guidance as to what the best approach is.

>>>>

One potential limitation is that the model is complex and has many parameters
in order to truly uncover the relevant patterns in the data. In order to
properly estimate these parameters at training time, a large amount of data
must be available to the training algorithm. The assumption that we have a
large amount of sequence data available for training on should generally
hold for most MOOC courses, but this assumption may be problematic if
attempting to apply our model to smaller online (or on-campus) courses.

@@

Our proposed model is flexible in the patterns it can discover in two main
ways. The first is that the granularity of the patterns can be adjusted by
changing the segmentation strategy one uses to divide the user action
stream into discrete ``sessions.'' The other is in the number of latent
states $K$ that are used to describe the segmented action sequences. One
drawback of these two lines of flexibility is that there is not necessarily
a clear answer as to the ``correct'' approach for both in any given
scenario. Varying the segmentation strategy changes the granularity of the
patterns that are required to be explained by the latent states, which will
change their meaning. Varying the number of latent states increases the
flexibility of the model, but also may result in latent states that are not
substantially different from the other latent states and/or latent states
with very low probability. This flexibility forces a user of our model to
make some assumptions about the patterns they wish to find (granularity,
diversity), and the model itself does not necessarily provide clear
guidance as to what the best approach is.

### 2017-06-14 sec-model.tex
Update paper for reviewer b; add response.

Our general idea is to use a probabilistic generative model to model 
the student activities as recorded in a MOOC log, which means we will 
assume that all the observed student activities are samples drawn (i.e., ``generated'') from 
a parameterized probabilistic model. We can then estimate the 
parameter values of the probabilistic model by fitting the model to a specific
MOOC log data set. The estimated parameter values could then be treated as 
the latent ``knowledge'' discovered from the data. Because such a generative model 
attempts to fit {\em all} the data, it enables us to discover interesting patterns
that can explain the {\em overall} behavior of a student or the {\em common} behavior patterns shared by many students.

An HMM  is a specific probabilistic generative model with a ``built-in'' state transition system
that would control the data to be generated by the model, thus it is especially 
suitable for modeling sequence data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would be in 
one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
another state stochastically according to a transition matrix that specifies the probability of
going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
The association of such a latent sequence of state transitions with the observed symbols makes
it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the
actual behavior of a student. We may say that students transition through different ``task states'' (or ``behavior states'')  in the process of study. 
One such task state may be to learn about a topic by mostly watching
lecture videos, another task state may be to work on quizzes, and yet
another may be to participate in forum discussions. While in each of these
different states, the student would tend to exhibit different surface
``micro'' behaviors. For example, in the lecture study state, the student
would tend to have many video-watching related behaviors and occasionally
forum activities, while in the quiz-taking state (in order to pass each
module), the student would tend to show many quiz-related ``micro''
activities as well as asking questions or checking discussions on the
forum. Note that due to the complexity of the student behavior, it is very
difficult to accurately {\em prescribe} the specific surface ``micro''
behavior patterns for each state in advance, especially without  prior knowledge about the students. For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations. 
The major motivations for using an HMM are that 
\begin{enumerate*}[label=(\arabic*)]
    \item it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main target to discover and characterize, and 
    \item it does not make any assumption about which latent task/behavior
      state must be associated with which observed activities or how a
      student would move from one state to another, but instead allows our
      data to ``tell'' us what kind of associations are most likely,
      what kind of transitions are most probable, and which states tend to be more long-lasting for any particular set of students. 
\end{enumerate*}

However, if we use an ordinary HMM to analyze our data, we would treat each
observed ``micro'' activity (e.g., video watching, or forum post reading)
as an output symbol, and thus the output distribution $p(x|u)$ for each
discovered latent state would be a simple distribution over all kinds of
observable micro activities recorded in our log data (e.g., 50\% lecture
watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading,
...). While such a distribution is meaningful and can already help us
interpret the corresponding latent state, it only gives us a rather
superficial characterization of student behavior.

Ideally, we want $p(x|u)$ to characterize the directly observable ``micro''
behavior in more detail to further capture the relations and dependencies
of these micro activities. To this end, we would treat an {\em entire
sequence} of micro activities (e.g., one session of activities) as an
observed ``symbol'' from a latent state, and further model the generation
of such a sequence with another Markov model where we treat each micro
activity as an {\em observable} state, and model the transitions between
these activity states in very much the same way as the state transitions in
HMM.

>>>>

Our general idea is to use a probabilistic generative model to model the
student activities as recorded in a MOOC log, which means we will assume
that all the observed student activities are samples drawn (i.e.,
``generated'') from a parameterized probabilistic model. We can then
estimate the parameter values of the probabilistic model by fitting the
model to a specific MOOC log data set. The estimated parameter values could
then be treated as the latent ``knowledge'' discovered from the data.
Because such a generative model attempts to fit \emph{all} the data, it
enables us to discover interesting patterns that can explain the
\emph{overall} behavior of a student or the \emph{common} behavior patterns
shared by many students.

An HMM is a specific probabilistic generative model with a ``built-in''
state transition system that would control the data to be generated by the
model, thus it is especially suitable for modeling sequence
data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would
be in one of $k$ states $U=\{u_1,\ldots,u_k\}$, and at the next moment, the
HMM would move to another state stochastically according to a transition
matrix that specifies the probability $p(u_i \mid u_j)$ of going to state
$u_i$ when the HMM is currently in state $u_j$. When the HMM is in state
$u$, the HMM can generate an observable data point $x$ according to an
output probabilistic model $p(x \mid u)$. Thus if we ``run'' an HMM for $N$
time points denoted by $t=1,\ldots,N$, the HMM could ``generate'' a sequence
of observations $x_1 \ldots x_N$, where each $x_i$ is an output symbol by
going through a sequence of \emph{hidden states} $w_1 \ldots w_N$ where
$w_i \in U$ is a state. The association of such a latent sequence of state
transitions with the observed symbols makes it possible to use the HMM to
``decode'' the latent behavior of students behind the surface behavior we
directly observe in the log data, allowing for understanding student
behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the
actual behavior of a student. We may say that students transition through
different ``task states'' (or ``behavior states'')  in the process of
study.  One such task state may be to learn about a topic by mostly
watching lecture videos, another task state may be to work on quizzes, and
yet another may be to participate in forum discussions. While in each of
these different states, the student would tend to exhibit different surface
``micro'' behaviors. For example, in the lecture study state, the student
would tend to have many video-watching related behaviors and occasionally
forum activities, while in the quiz-taking state (in order to pass each
module), the student would tend to show many quiz-related ``micro''
activities as well as asking questions or checking discussions on the
forum. Note that due to the complexity of the student behavior, it is very
difficult to accurately \emph{prescribe} the specific surface ``micro''
behavior patterns for each state in advance, especially without  prior
knowledge about the students. For example, forum activities are likely
interleaved with other activities   in every task state and the
interleaving pattern can be somewhat irregular with potentially many
variations.  The major motivations for using an HMM are that
\begin{enumerate*}[label=(\arabic*)]
  \item it uses a probabilistic model (the output probability distribution
    $p(x \mid u)$ conditioned on each state) to directly capture the
    inevitable uncertainty in the association of surface ``micro''
    activities with their corresponding latent task/behavior state, which
    is often our main target to discover and characterize, and
  \item it does not make any assumption about which latent task/behavior
    state must be associated with which observed activities or how a
    student would move from one state to another, but instead allows our
    data to ``tell'' us what kind of associations are most likely, what
    kind of transitions are most probable, and which states tend to be more
    long-lasting for any particular set of students.
\end{enumerate*}

However, if we use an ordinary HMM to analyze our data, we would treat each
observed ``micro'' activity (such as video watching or forum post reading)
as an output symbol, and thus the output distribution $p(x \mid u)$ for each
discovered latent state would be a simple distribution over all kinds of
observable micro activities recorded in our log data (e.g., 50\% lecture
watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading
\ldots). While such a distribution is meaningful and can already help us
interpret the corresponding latent state, it only gives us a rather
superficial characterization of student behavior.

Ideally, we want $p(x \mid u)$ to characterize the directly observable
``micro'' behavior in more detail to further capture the relations and
dependencies of these ``micro'' activities. To this end, we would treat an
\emph{entire sequence} of ``micro'' activities (e.g., one session of
activities) as an observed ``symbol'' from a latent state, and further
model the generation of such a sequence with another Markov model where we
treat each micro activity as an \emph{observable} state, and model the
transitions between these activity states in very much the same way as the
state transitions in an HMM.

### 2017-06-14 sec-related-work.tex
Update paper for reviewer b; add response.

\citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
with a set of actions that is defined similarly to ours. However, their
method for finding the common behavioral patterns involves a manual
clustering step to identify behavioral ``motifs'', which is then followed
by an automatic (rule-based) assignment of all sequences to these motifs.
Our method, by contrast, attempts to do this automatically: the latent
state representations obtained by our model attempt to capture similar
meanings to their behavior motifs in a completely automated fashion. They
also automatically generate and investigate Markov models for different
MOOCs, but do so by considering \emph{all} student action sequences as
belonging to a \emph{single Markov model}. In our approach, we allow each
student behavior sequence to belong to one of $K$ different Markov models
(and further model the transition probabilities between these latent state
Markov models between each sequence a student generates). Thus, their
Markov models presented are a special case of our model when $K = 1$.

\citet{Faucon:2016:EDM} proposed a semi-Markov model for simulating MOOC
students. They produce behavior profiles that characterize groups of
students in the form of semi-Markov models like our proposed model does,
but they differ from our work in two major ways. First, they assume that a
student can belong to only one behavior profile across the entire course
rather than allowing this profile to change over time.  Because we do not
have this restriction, our model is also able to learn the transition
probabilities between the different behavior profiles we discover.

>>>>

\citet{Davis:2016:EDM} investigate frequent student behavior pattern chains
with a set of actions that is defined similarly to ours. However, their
method for finding the common behavioral patterns involves a manual
clustering step to identify behavioral ``motifs,'' which is then followed
by an automatic (rule-based) assignment of all sequences to these motifs.
Our method, by contrast, attempts to do this automatically: the latent
state representations obtained by our model attempt to capture similar
meanings to their behavior motifs in a completely automated fashion. They
also automatically generate and investigate Markov models for different
MOOCs, but do so by considering \emph{all} student action sequences as
belonging to a \emph{single Markov model}. In our approach, we allow each
student behavior sequence to belong to one of $K$ different Markov models
(and further model the transition probabilities between these latent state
Markov models between each sequence a student generates). Thus, their
Markov models presented are a special case of our model when $K = 1$.

\citet{Faucon:2016:EDM} proposed a semi-Markov model for simulating MOOC
students. They produce behavior profiles that characterize groups of
students in the form of semi-Markov models like our proposed model does,
but they assume that a student can belong to only one behavior profile
across the entire course rather than allowing this profile to change over
time. Because we do not have this restriction, our model is also able to
learn the transition probabilities between the different behavior profiles
we discover.
